---
title: "Introduction to combreg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to combreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7)
has_lpsolve <- requireNamespace("lpSolve", quietly = TRUE)
```

## The model

`combreg` fits Bayesian regression models whose response is a binary vector
constrained to a polytope,

$$\mathcal{Y} = \{\, y \in \{0,1\}^d : A y \le b \,\},$$

with $A$ an integer, totally unimodular constraint matrix. Examples include
matchings, assignments, and other combinatorial structures. Each observation
has a latent Gaussian utility

$$\zeta_i = \beta^\top x_i + \varepsilon_i, \qquad
\varepsilon_i \sim N(0, I_d),$$

and the observed response is the feasible utility maximizer

$$y_i = \arg\max_{y \in \mathcal{Y}} \; \zeta_i^\top y.$$

Posterior sampling uses a Metropolis–Hastings-within-Gibbs scheme on an
augmented posterior in which each observation carries a dual certificate
$u_i \ge 0$ of the optimality of $y_i$; the dual updates use hit-and-run
sampling implemented in C++.

## Defining constraints

Constraints are declared once and validated (including an exhaustive total
unimodularity check for small systems):

```{r}
library(combreg)

A <- rbind(
  c(1, 1, 0),
  c(0, 1, 1)
)
con <- crr_constraints(A, b = c(1, 1))
con
```

`is_tum()` and `is_feasible()` are available as standalone utilities.

## Simulating data and fitting

`simulate_crr()` draws covariates, coefficients, and utilities, and maps each
utility vector to its constrained maximizer by integer programming (this
requires the `lpSolve` package):

```{r, eval = has_lpsolve}
sim <- simulate_crr(n = 300, p = 2, constraints = con, seed = 1)
head(sim$Y)
```

```{r, eval = has_lpsolve}
fit <- crr(sim$Y, sim$X, con,
           kernel = "exponential",
           n_iter = 1500, warmup = 500, seed = 1)
fit
```

Posterior summaries and diagnostics:

```{r, eval = has_lpsolve}
head(summary(fit))
plot(fit, pars = c("beta[1,1]", "beta[2,1]"))
```

Draws convert directly to the `coda` ecosystem via `coda::as.mcmc(fit)`, and
to `posterior::draws_array` via `posterior::as_draws(fit)`.

## Why constraints matter

Ignoring the constraint structure biases estimation. The `"unconstrained"`
method fits independent Albert–Chib probit regressions per coordinate — a
natural but wrong model for constrained responses:

```{r, eval = has_lpsolve}
fit_un <- crr(sim$Y, sim$X, method = "unconstrained",
              n_iter = 1500, warmup = 500, seed = 1)

rmse <- function(est, truth) sqrt(mean((est - truth)^2))
c(constrained   = rmse(coef(fit), sim$beta),
  unconstrained = rmse(coef(fit_un), sim$beta))
```

The constrained sampler recovers $\beta$ markedly better because the
likelihood correctly conditions on $y_i$ being the feasible maximizer rather
than a vector of independent signs.

## Building custom models from primitives

Models with structured (non-linear-in-$X$) means can reuse the exported
primitives — `init_dual()`, `sample_dual()`, `draw_utility()`,
`sample_utility()`, `dual_feasible()`, `coef_precompute()`, `update_coef()` —
and supply their own coefficient updates, so that no constraint logic needs
to be reimplemented. See their reference pages for the exact conditional
distributions.
