---
title: "Introduction to combreg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to combreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7)
has_lpsolve <- requireNamespace("lpSolve", quietly = TRUE)
```

## The model

`combreg` fits Bayesian regression models whose response is a binary vector
constrained to a polytope,

$$\mathcal{Y} = \{\, y \in \{0,1\}^d : A y \le b \,\},$$

with $A$ an integer, totally unimodular constraint matrix. Examples include
matchings, assignments, and other combinatorial structures. Each observation
has a latent Gaussian utility

$$\zeta_i = \beta^\top x_i + \varepsilon_i, \qquad
\varepsilon_i \sim N(0, I_d),$$

and the observed response is the feasible utility maximizer

$$y_i = \arg\max_{y \in \mathcal{Y}} \; \zeta_i^\top y.$$

Posterior sampling uses a Metropolis–Hastings-within-Gibbs scheme on an
augmented posterior in which each observation carries a dual certificate
$u_i \ge 0$ of the optimality of $y_i$; the dual updates use hit-and-run
sampling implemented in C++.

## Defining constraints

Constraints are declared once and validated (including an exhaustive total
unimodularity check for small systems):

```{r}
library(combreg)

A <- rbind(
  c(1, 1, 0),
  c(0, 1, 1)
)
con <- crr_constraints(A, b = c(1, 1))
con
```

Here `A` and `b` describe the polytope $\{\, y \in \{0,1\}^d : A y \le b \,\}$
directly: each row of `A y <= b` is one linear inequality carving the feasible
set out of the binary cube. The two rows above encode "at most one of
coordinates 1--2" and "at most one of coordinates 2--3".

`is_tum()` and `is_feasible()` are available as standalone utilities. Total
unimodularity of `A` is what makes the dual-certificate augmentation exact, so
it is checked at construction; you can also test a matrix directly:

```{r}
is_tum(A)                                     # TRUE: this A is TUM
is_tum(rbind(c(1, 1, 0), c(1, 0, 1), c(0, 1, 1)))  # FALSE (odd cycle)
```

`is_feasible()` checks response rows against the polytope --- each returned
logical answers "is this row in $\mathcal{Y}$?":

```{r}
candidates <- rbind(
  c(1, 0, 1),   # A y = (1, 1) <= (1, 1): feasible
  c(1, 1, 0)    # A y = (2, 1): violates row 1
)
is_feasible(con, candidates)
```

The same pattern generalizes to any TUM system. A single "at most one active
coordinate" simplex, for instance, is one dense row:

```{r}
d <- 3
A_simplex <- matrix(1, nrow = 1, ncol = d)
con_simplex <- crr_constraints(A_simplex, b = 1)
con_simplex
is_feasible(con_simplex, rbind(diag(d)[1, ], rep(1, d)))  # e_1 ok, all-ones not
```

## Simulating data and fitting

`simulate_crr()` draws covariates, coefficients, and utilities, and maps each
utility vector to its constrained maximizer by integer programming (this
requires the `lpSolve` package):

```{r, eval = has_lpsolve}
sim <- simulate_crr(n = 300, p = 2, constraints = con, seed = 1)
head(sim$Y)
```

```{r, eval = has_lpsolve}
fit <- crr(sim$Y, sim$X, con,
           kernel = "exponential",
           n_iter = 1500, warmup = 500, seed = 1)
fit
```

Posterior summaries and diagnostics:

```{r, eval = has_lpsolve}
head(summary(fit))
plot(fit, pars = c("beta[1,1]", "beta[2,1]"))
```

Draws convert directly to the `coda` ecosystem via `coda::as.mcmc(fit)`, and
to `posterior::draws_array` via `posterior::as_draws(fit)`. For the full
diagnostics workflow — the structured `crr_diagnostics()` report, violin /
ESS / efficiency / residual plots, and benchmarking with `crr_benchmark()` —
see `vignette("diagnostics")`.

## Why constraints matter

Ignoring the constraint structure biases estimation. The `"unconstrained"`
method fits independent Albert–Chib probit regressions per coordinate — a
natural but wrong model for constrained responses:

```{r, eval = has_lpsolve}
fit_un <- crr(sim$Y, sim$X, method = "unconstrained",
              n_iter = 1500, warmup = 500, seed = 1)

rmse <- function(est, truth) sqrt(mean((est - truth)^2))
c(constrained   = rmse(coef(fit), sim$beta),
  unconstrained = rmse(coef(fit_un), sim$beta))
```

The constrained sampler recovers $\beta$ markedly better because the
likelihood correctly conditions on $y_i$ being the feasible maximizer rather
than a vector of independent signs.

## The adaptive `zeta_block` controller

The latent-utility MH update does not refresh all $d$ coordinates at once.
Each sweep resamples a random sub-block of the coordinates, and the block
size --- `zeta_block` --- controls the trade-off between per-step acceptance
and how far the utilities move: small blocks accept often but mix slowly,
large blocks propose bolder moves that are rejected more often. The optimal
size depends on the constraint geometry, not just on $d$, so a good default
cannot be a fixed number.

The default `zeta_block = "adaptive"` handles this automatically. During
warmup a dual-averaging controller tunes the block size to hit
`zeta_target_accept` (default `0.6`), then freezes it for the sampling phase
so the proposal is fixed (valid MCMC). The tuned value is reported by
`print()` and stored on the fit; the `fit` above already used it:

```{r, eval = has_lpsolve}
fit$zeta_block_tuned
```

You therefore never need to tune this by hand. When exact reproducibility of
the proposal across runs matters --- e.g. regression tests --- pin an integer
instead, which disables adaptation:

```{r, eval = has_lpsolve}
fit_fixed <- crr(sim$Y, sim$X, con,
                 n_iter = 800, warmup = 300, seed = 1,
                 control = crr_control(zeta_block = 100))
fit_fixed$zeta_block_tuned   # block sizes are capped at d
```

## Customizing the sampler

Every knob is set through `crr()` and `crr_control()`. A fully customized
fit:

```{r, eval = has_lpsolve}
fit_custom <- crr(sim$Y, sim$X, con,
                  kernel = "half_gaussian",         # half-Gaussian dual kernel
                  prior  = crr_prior(sd = 10),      # wider N(0, 100) prior
                  n_iter = 1000, warmup = 400,      # total / discarded sweeps
                  thin   = 2,                       # keep every 2nd draw
                  chains = 2,                       # independent chains (serial)
                  seed   = 1,                        # reproducible RNG stream
                  control = crr_control(
                    n_iter_hit_and_run = 50,        # inner hit-and-run steps
                    zeta_block = "adaptive",        # auto-tune the block size
                    n_threads = 1))                 # OpenMP threads (result-invariant)
fit_custom
```

Running multiple chains lets you assess convergence with the usual
between-chain diagnostics (see `vignette("diagnostics")`).

## Formula interface

`X` may be a one-sided formula evaluated against a data frame, which is
convenient when covariates live in columns. Because `simulate_crr()` uses a
no-intercept design, drop the intercept with `~ 0 + ...` so the design matrix
matches the simulated one:

```{r, eval = has_lpsolve}
df <- data.frame(x1 = sim$X[, 1], x2 = sim$X[, 2])
fit_formula <- crr(sim$Y, ~ 0 + x1 + x2, con, data = df,
                   n_iter = 800, warmup = 300, seed = 1)
coef(fit_formula)
```

The formula path uses `model.matrix()`, so factor expansion and interactions
work as in any R modelling function.

## Building custom models from primitives

Models with structured (non-linear-in-$X$) means can reuse the exported
primitives — `init_dual()`, `sample_dual()`, `draw_utility()`,
`sample_utility()`, `dual_feasible()`, `coef_precompute()`, `update_coef()` —
and supply their own coefficient updates, so that no constraint logic needs
to be reimplemented. See their reference pages for the exact conditional
distributions.
