---
title: "Diagnostics and benchmarking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5)
has_lpsolve <- requireNamespace("lpSolve", quietly = TRUE)
knitr::opts_chunk$set(eval = has_lpsolve)
```

This vignette shows the diagnostic and benchmarking workflow for `crr()`
fits: the structured diagnostics report, the plotting methods, and the
one-call benchmark against the unconstrained baseline. It mirrors, at a
CRAN-friendly scale, the simulation comparisons in *Statistical Modeling of
Combinatorial Response Data*.

## Setup: simulate and fit

```{r}
library(combreg)

A <- rbind(
  c(1, 1, 0),
  c(0, 1, 1)
)
con <- crr_constraints(A, b = c(1, 1))
sim <- simulate_crr(n = 300, p = 2, constraints = con, seed = 42)

fit <- crr(sim$Y, sim$X, con,
           n_iter = 4000, warmup = 2000, chains = 2, seed = 42)
fit
```

## The diagnostics report

`crr_diagnostics()` assembles everything needed to judge a run in one
object: per-coefficient posterior summaries, effective sample sizes (total
and per second of sampling), split-Rhat, MH acceptance, in-sample regression
fit, and posterior predictive goodness-of-fit checks. In simulation the true
coefficients are known, so passing `beta` additionally reports estimation
error (RMSE) and posterior-interval coverage. Potential problems (Rhat above
1.05, ESS below 100, very low acceptance, extreme predictive p-values) are
collected as warnings.

```{r}
report <- crr_diagnostics(fit, beta = sim$beta)
report
```

The underlying per-parameter table is available as `report$table`, and the
predictive-check object as `report$ppc`.

### Reading the goodness-of-fit checks

The raw exact-match and accuracy rates are *not* by themselves evidence for
or against the model: the response is the maximizer of a noisy utility, so
even the true $\beta$ cannot predict $y_i$ perfectly — point-prediction
accuracy is bounded by the intrinsic noise. The posterior predictive checks
calibrate this. Each replicate re-simulates responses from the fitted model
and recomputes the fit statistics; the model is adequate when the observed
statistics are typical of that distribution (two-sided p-values away from
0). Here the observed exact-match rate sits in the middle of its predictive
distribution — the seemingly modest rate is exactly what a *correct* model
predicts for this noise level, and the RMSE and coverage against the known
truth confirm accurate estimation. `crr_ppc()` runs the checks standalone
with more replicates if needed.

## Plotting methods

All plots operate on the post-warmup, thinned draws.

Trace and autocorrelation:

```{r}
plot(fit, type = "trace", pars = c("beta[1,1]", "beta[2,1]"))
plot(fit, type = "acf", pars = c("beta[1,1]", "beta[2,1]"))
```

Marginal posteriors of all coefficients as violins (median dot, central 95%
interval bar):

```{r}
plot(fit, type = "violin")
```

Effective sample sizes and sampling efficiency (ESS per second of sampling
time, the efficiency measure used for method comparison in the paper):

```{r}
plot(fit, type = "ess")
plot(fit, type = "ess_time")
```

Residual heat map of the posterior point-estimate fit
(`Y - fitted(fit)`, entries in $\{-1, 0, 1\}$):

```{r, fig.height = 5.5}
plot(fit, type = "residual")
```

When the truth is known, the coefficient estimation error
$\hat\beta - \beta$ as a heat map:

```{r}
plot(fit, type = "beta_diff", beta = sim$beta)
```

For anything beyond these built-ins, convert the draws with
`coda::as.mcmc(fit)` or `posterior::as_draws(fit)` and use those ecosystems
directly.

## Benchmarking against the unconstrained baseline

`crr_benchmark()` fits several methods to the same data with identical
settings and reports timing, efficiency, in-sample fit, and — when the true
coefficients are known, as in simulation — RMSE and posterior-interval
coverage:

```{r}
bm <- crr_benchmark(sim$Y, sim$X, con, beta = sim$beta,
                    n_iter = 2000, warmup = 1000, seed = 42)
bm
```

Two effects visible here reproduce the paper's findings: the constrained
sampler estimates $\beta$ with smaller error and better-calibrated
intervals, because the unconstrained probit baseline conditions on each
coordinate independently and is biased when responses are in fact
constrained maximizers.

The fitted models are kept in `bm$fits`, so any of the diagnostics above can
be applied to each method afterwards, e.g.
`plot(bm$fits$unconstrained, type = "violin")`.
